---
title: "Analysis of Urban Ministries of Durham Data"
author: "Joseph Foster"
date: "10/07/2019"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Motivation of Study

Urban Ministries of Durham (UMD) is helping to end homelessness through the distribution of food, providing shelter, and helping develop futures. Here I aim to assess how well the organization has distributed food in the past, and what factors may have influenced that distribution.

### Report


I first looked at how much food was distributed annually from 2006 to current date. What I found was the total annual food provided has increased over the last 5 years, and has been trending upward since 2006. Note data collection from the current year 2019 is likely incomplete and is probably driving the reported reduction of food provided to clients. Nonetheless, these data may be helpful for UMD when pitching to new funding sources, as they can show the the organization's ability to provide more food to clients year-to-year.

```{r fig1, fig.height = 5, fig.width = 8, fig.align = "center", echo=FALSE, message=FALSE}
library(tidyverse)
#Read in UMD data
umd_df = read.csv('../data/UMD_Services_Provided_20190719.tsv',sep = '\t',header = TRUE)

#Data cleaning
umd_df_clean = umd_df %>%  
  #Format data variable
  separate(Date, sep = "/", into = c("Month", "Day", "Year")) %>%
  #Reduce the dataframe to specific year range and remove Food.Pounds outliers 
  filter(Year >= 2006 & Year <= 2019 & Food.Pounds < 100)

#What is the total weight of food provided per year?
 umd_df_clean%>%
  #Reduce dataframe to only include two variables
  select(Year,Food.Pounds) %>%
  #Remove any row with NaN
  drop_na(Food.Pounds) %>%
  #Group data by year
  group_by(Year) %>%
  #Find aggregate food provide per year
  summarise(food.give = sum(Food.Pounds)) %>% 
  #Create bar plot
  ggplot(aes(x = Year, y = food.give)) +
  geom_bar(stat = "identity") +
  labs(x = "Year",
       y = "Aggregate Food distributed (lbs.)",
       face = 'bold') +
  theme_minimal()

```


Given the upward trend in total food distribution over the last decade I next looked to see how that food was distributed on a client-by-client basis. To understand how much food was being distributed per transaction, I looked at the distribution of food received during each interaction for every year over the past 13 years. The average amount of food given at each distribution has been stable for nearly a decade. This analysis also suggests UMD has done well at reducing any outlier food allotments since 2011. Taken together these data may suggest the operation procedures driving food distribution are helping to create an equality in the distribution of food.

```{r fig2, fig.height = 5, fig.width = 10, fig.align = "center", echo=FALSE, message=FALSE}
#What is the average amount of food given per interaction in a given year?
umd_df_clean %>%
  #Reduce dataframe to only include two variables
  select(Year,Food.Pounds) %>%
  #Remvove rows with NaN
  drop_na(Food.Pounds) %>%
  #Create box plot
  ggplot(aes(x = Year, y = Food.Pounds)) +
  geom_violin() +
  geom_boxplot(width=0.1) +
  labs(x = "Year",
       y = "Food provided at each transaction (lbs.)",
       face = 'bold') +
  theme_minimal()
```


I next sought to determine if other variables impacted how UMD is distributing food. Specifically, to understand if paying a bill impacts the amount of food received per interaction, I looked at the amount of food received per interaction on an annual basis for individuals who reported paying a bill and those who did not. Over the past 13 years, only two years showed a significant difference (2006, 2008) in the amount of food received when the client also reported paying a bill (Wilcoxn rank sum test, ns: p > 0.05, * p <= 0.05, ** p <= 0.01). However, these data shouldn't be over interpreted as there were only 127 interactions where a bill was reported as paid, while there were over 65,000 interactions included in this analysis with no mention of paying bills.

```{r fig, fig.height = 5, fig.width = 8, fig.align = "center", echo=FALSE, message=FALSE}
library(ggpubr)
#Overall distribution of food given out by year
overall.food <- umd_df %>%  
  separate(Date, sep = "/", into = c("Month", "Day", "Year")) %>%
  filter(Year >= 2006 & Year <= 2019, Food.Pounds < 100) %>%
  select(Year,Type.of.Bill.Paid,Food.Pounds) %>%
  drop_na(Food.Pounds)
#Add label for plotting
overall.food$bill.status <- "no_bill_info"

#The amout of food given to individual paying bills
paid.bill.food <-umd_df %>%  
  #Format date variable
  separate(Date, sep = "/", into = c("Month", "Day", "Year")) %>%
  #Reduce the dataframe to specific year range and remove Food.Pounds outliers 
  filter(Year >= 2006 & Year <= 2019 & Food.Pounds < 100) %>%
  #Reduce dataframe to 3 variables
  select(Year,Type.of.Bill.Paid,Food.Pounds) %>%
  #Remove rows with NaN
  drop_na(Type.of.Bill.Paid) %>%
  drop_na(Food.Pounds) %>%
  #Remove rows with nothing in cell
  filter(Type.of.Bill.Paid != "")

#Add label for plotting
paid.bill.food$bill.status <- "paid_bill"

#Concatenate two dataframes
merge <- rbind(overall.food,paid.bill.food)

merge %>% 
  #Focus on years after 2006
  filter(Year >= 2006) %>%
  #Box Plot
  ggplot(aes(x = Year, y = Food.Pounds, fill=bill.status)) +
  geom_boxplot() +
  theme_minimal() +
  labs(y = "Food provided at each transaction (lbs.)",
       face = "bold") + 
  #Determine if difference of each pairing is statistically significant (Wilcox)
  stat_compare_means( aes(label = ..p.signif..), 
                      label.x = 1.5, label.y = 100)
```


### Conclusion
In summary, over the past decade UMD has been able to increase the total amount of food reaching its clients, while also holding stable the average amount of food given per interaction. Although the data for clients paying bills was sparse, there is some evidence to suggest that reporting payment of bills doesn't have an impact on the average amount of food clients receive at a given interaction. Given, one of the central tenants of UMD is to provide food to the community these results are encouraging. 


### Notes on future data collection
1. While cleaning the data I noticed there were lines dating back to 1931, which predates the existence of UMD. There are also records listing future dates. It would be helpful to understand if these data are erroneous or if they have some other type of significance. 

2. While looking into the total amount of food provided each year I came across a record (client #: 12943) that had received 450121 pounds of food during one interaction. I assume that this was a data entry error. It would be helpful to know what the upper and lower limits are on the amount of food distributed at any one time. 

3. The analyses reported here focused on the year range 2006-2019 because the data before 2006 was sparse in comparison. Are there additional records from earlier years?

